Mergeable Types

### 2017-02-28 paper/related.tex

Our idea of versioning state bears resemblance to Concurrent
Revisions~\cite{BBL+10}, a programming abstraction that provides
deterministic concurrent execution.  The idea of using revisions as a
means to programming eventually consistent distributed systems was
further developed in~\cite{BFL+12,BLF+12}.  The
\name\ programming model, however, differs substantially from a
concurrent revisions model because it imposes no distinction between
\emph{servers}, machines that hold global state, and \emph{clients},
devices that operate on local, potentially stale, data - any
computation executing in a distributed environment is free to fork new
versions, and synchronize against other replicated state.  Our model,
furthermore, supports fully decentralized operation and is robust to
network partitions and failures. Just as significantly, \name\ allows
applications to customize join semantics with programmable merge
operations.  Indeed, the integration of a version-based mechanism
within OCaml allows a degree of type safety, composability, and
profitable use of polymorphism not available in related systems.

@@

There have been numerous proposals over the years for realizing
distributed functionality in a functional programming context.
Poly/ML~\cite{Mat97} and Facile~\cite{TLK+93} extend Standard ML,
while Acute~\cite{SLW+05} and HashCaml~\cite{BSS+06} extend OCaml, with
distributed programming abstractions and support, with particular
focus on type-safe marshalling and mobility.  Kali-Scheme~\cite{CJK95}
defines a distributed extension for Scheme-48, and Cloud
Haskell~\cite{EBPJ11} describes a domain-specific language that
enables development of distributed Haskell programs.
Obliq~\cite{Car95} explores issues related to mobility, lexical
scoping, and network references in an object-oriented framework that
enables safe (i.e., lexical) transmission of closures.  However,
unlike \name, none of these systems consider issues of object
replication or consistency as central to their programming models.
The realities of modern-day scalable distributed systems, especially
those that are constructed from geographically distributed
participants, dictate that we must necessarily grapple with
consistency issues to achieve any kind of reasonable performance; it
is noteworthy that replication is inherent in many widely-used
cloud-based distributed system implementations today (e.g.,
Cassandra~\cite{Cassandra} or Dynamo~\cite{Dynamo}), which provide
only weak consistency guarantees.  Rather than exposing a low-level
operational treatment of replication, foisting the burden of reasoning
about replicated state onto the programmer, our primary contribution
is the development of a principled high-level approach to framing
these issues that abstracts these kinds of system-level details.  Our
goal is to leverage functional programming principles to minimize
disruption to the way programmers structure and reason about their
applications, without compromising efficiency or performance.  Indeed,
the semantics of mergeable types and versioning has utility in any
concurrent setting that must deal with non-trivial coordination and
synchronization costs, even without taking distribution or replication
issues into account.

>>>>

Our idea of versioning state bears resemblance to Concurrent
Revisions~\cite{BBL+10}, a programming abstraction that provides
deterministic concurrent execution.  The idea of using revisions as a
means to programming eventually consistent distributed systems was
further developed in~\cite{BFL+12}.  The
\name\ programming model, however, differs substantially from a
concurrent revisions model because it imposes no distinction between
\emph{servers}, machines that hold global state, and \emph{clients},
devices that operate on local, potentially stale, data - any
computation executing in a distributed environment is free to fork new
versions, and synchronize against other replicated state.  Our model,
furthermore, supports fully decentralized operation and is robust to
network partitions and failures. Just as significantly, \name\ allows
applications to customize join semantics with programmable merge
operations.  Indeed, the integration of a version-based mechanism
within OCaml allows a degree of type safety, composability, and
profitable use of polymorphism not available in related systems.

@@

There have been numerous proposals over the years for realizing distributed
functionality in a functional programming context. Poly/ML~\cite{Mat97} and
Facile~\cite{TLK+93} extend Standard ML, while Acute~\cite{SLW+05} and
HashCaml~\cite{BSS+06} extend OCaml, with distributed programming abstractions
and support, with particular focus on type-safe marshalling and mobility.
Kali-Scheme~\cite{CJK95} defines a distributed extension for Scheme-48, and
Cloud Haskell~\cite{EBPJ11} describes a domain-specific language that enables
development of distributed Haskell programs. Obliq~\cite{Car95} explores issues
related to mobility, lexical scoping, and network references in an
object-oriented framework that enables safe (i.e., lexical) transmission of
closures. Orleans~\cite{Orleans} provides persistent actor-based programming
model for elastic cloud application, but the focus is on strong consistency,
mirroring a single-threaded execution.

However, unlike \name, none of these systems consider issues of object
replication or consistency as central to their programming models. The
realities of modern-day scalable distributed systems, especially those that are
constructed from geographically distributed participants, dictate that we must
necessarily grapple with consistency issues to achieve any kind of reasonable
performance; it is noteworthy that replication is inherent in many widely-used
cloud-based distributed system implementations today (e.g.,
Cassandra~\cite{Cassandra} or Dynamo~\cite{Dynamo}), which provide only weak
consistency guarantees.  Rather than exposing a low-level operational treatment
of replication, foisting the burden of reasoning about replicated state onto
the programmer, our primary contribution is the development of a principled
high-level approach to framing these issues that abstracts these kinds of
system-level details.  Our goal is to leverage functional programming
principles to minimize disruption to the way programmers structure and reason
about their applications, without compromising efficiency or performance.
Indeed, the semantics of mergeable types and versioning has utility in any
concurrent setting that must deal with non-trivial coordination and
synchronization costs, even without taking distribution or replication issues
into account.

\name shares resemblance to conflict-free replicated data types
(CRDT)~\cite{crdt}. CRDTs define abstract data types such as counters, sets,
etc., with commutative operations such that the state of the data type always
converges. Unlike CRDTs, the operations on mergeable types in \name need not
commute and the reconciliation protocol is defined by user-defined merge
functions. Moreover, CRDTs are not composable and each CRDT tends to be built
from scratch over the network protocol. Compare to this, \name types are
composable; mergeable ropes are polymorphic over its mergeable content. \name
uses 3-way merges using the lowest common ancestor, which is critical for all
of our user-defined merges. However, CRDTs do not have the benefit of lowest
common ancestor for merges and are only presented with the two concurrent
versions. If a 3-way merge is desired, then the causal history has to be
explicitly encoded in the data type. As a result, constructing even simple data
types like counters are more complicated using CRDTs~\cite{crdt} compared to
their implementation in \name.
